---
title: A Personal Pandas, Dplyr, Data.table Repository
author: Yang Wu
date: '2022-01-08'
slug: a-personal-pandas-dplyr-data-table-repository
categories:
  - Python
  - R
tags:
  - Data Wrangling
  - R Programming
  - Python Programming
subtitle: ''
summary: 'A running list of everyday data wrangling code snippets (Updated irregularly)'
authors: []
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  blogdown::html_page:
    toc: true
---

```{r global_options, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(results = "hold")
library(tibble)
library(dplyr)
library(data.table)
library(reticulate)
reticulate::use_condaenv(
  condaenv = "python_automation",
  conda = "/Users/kenwu/opt/anaconda3/envs/python_automation",
  required = TRUE
)
```

# Motivation

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; The content and formatting of this post is *heavily* inspired by the following comprehensive and awesome [post](https://atrebas.github.io/post/2019-03-03-datatable-dplyr/) by [Atrebas](https://atrebas.github.io/about/). You could also find him on Twitter [here](https://twitter.com/atrebas). I use both `R` and `Python` extensively in my work everyday, especially the data wrangling packages and library `data.table`, `dplyr` (its related siblingz `dbplyr`, `dtplyr` etc.), and `pandas`. I find that, in my case, I often have to translate between different syntax and programmatic thinking when switching from one another. In `R`, there are ways to accomplishing tasks and there are more Pythonic ways to do things in `Python`. This is an attempt to create a list of code snippets that can hopefully become a reference, which will be updated over time, for myself and perhaps others as well.

# Packages and Library

For `R`:

```{r, eval=FALSE}
library(tibble)
library(dplyr)
library(data.table)
```

For `Python`:

```{python}
import pandas as pd
import numpy as np
```

# Data Frame Creation

## Dplyr

```{r}
tb <- tibble(
  x = c(1, 2, 3),
  y = c(2, 3, 4)
)
tb
```

## Data.table

```{r}
dt <- data.table(
  x = c(1, 2, 3),
  y = c(2, 3, 4)
)
dt
```

## Pandas

```{python}
df = pd.DataFrame({
  'x': [1, 2, 3], 
  'y': [2, 3, 4]
})
df
```







